---
title: "Data Science Applied to Ag - Final Project - Design"
author: "Muhammad Usman Khalid"
format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
    number-sections: true
    theme: cerulean
---

# Instructions  
This file contains both the **instructions** for the mid-term project and placeholders for your code. You are required to use this file to produce code, output, and answers to the questions below.  

Besides simply creating output, make sure to interpret the output. You will need to create tables and/or plots to arrive at the answers, and then comment on what you found below it.    

To get you setup, you will need to:  

  - Student #1: create a repository on your GitHub account. You can call this repository "2025dsa_finalproject_groupX", **where X is the number of your group**. Make it public, add a README, add a .gitignore for R.  
  - Student #1: follow the steps we did in class to start a new RStudio project with version control.  
  - Student #1: in your computer, create the sub-folders code, data, output, and move your data set into the `data` folder. Also, student 1 moves this current script into the `code` folder. Do a git stage, commit, push.  
  - Student #1: on GitHub, go the repository settings and invite your partner to be a collaborator in the repository. That will give them push permission.  
  - Now, both students should clone this repository on their computers like we did in class. Make sure this step works well and that you can pull and push from GitHub.  
  - Student 2, after cloning, does a git pull to get all these updates on their computer.  
  - Student 1 and 2 work together to update the README file. README files should explain what the repository is about, the goals of that project, who is working in it, and any other important details you may find.  
  
# Introduction  
Describe here the introduction of your problem. Consider this as a shortened version of your paper, where you will briefly discuss in 3-4 paragraphs what is the issue/gap in literature, and how the data you collected will help answer this gap.
With the rising global population, the demand for food production continues to increase. This has led to intensive agricultural practices, including the excessive use of nitrogen-based fertilizers to boost crop yields. While these fertilizers support productivity, they also pose environmental risks, especially when nitrate (NO₃⁻) leaches into water bodies, leading to groundwater contamination and eutrophication.

Nitrate is highly mobile in soil and can be easily lost through leaching or denitrification. Improving nitrogen retention and use efficiency is therefore a critical focus of sustainable agriculture. One promising approach is the use of Locally Effective Microorganisms (LEM) that may enhance nitrate availability through mineralization while potentially reducing losses via denitrification or leaching.

Although many studies have explored the use of microbial inoculants, limited work has been done to evaluate LEM application rates and their effects on nitrate concentration in organically managed soils. This study aims to fill this gap by evaluating the role of LEM in nitrate dynamics under corn production.

# Hypothesis and objectives  
Describe here your hypothesis, followed by your objectives. Make sure your hypothesis are testable and bold, and objectives are clear.  

Hypothesis:
Locally effective microorganisms will increase soil nitrate concentration by enhancing nitrogen mineralization.
Objective:
To determine the effect of LEM application rates on soil nitrate concentration and to identify the optimal LEM rate that maximizes nitrate availability without excessive nitrogen losses.

# Material and Methods  
Describe here your overall material and methods as it pertains to the analysis you will conduct, including study description, site/setup description, what equipment was used, etc. just like you would in a paper. Make sure to clearly explain what was measured and how.

Site & Microbial Inoculum Preparation:
The study was conducted at the Horticulture Farm in Watkinsville, Georgia in 2023. The field had been under organic management for 10 years. Solid LEM was prepared using forest litter from hardwood and pine forests, mixed with semolina (50 lbs), yeast, crushed charcoal, molasses (1 gallon), unpasteurized goat milk (1 gallon), and non-chlorinated water. After 4 weeks of fermentation, 3 kg of solid LEM was suspended in 8 gallons of water + 1 gallon molasses to make liquid LEM, fermented for 3 more weeks.

Planting and Application:
Corn (Wapsie Valley) was planted on May 15, 2022. LEM was applied at three rates: 0, 1.8, and 3.6 L/m² around corn stems. Soil samples (0–15 cm) were collected 4 weeks after application for nitrate (NO₃⁻) analysis.

## Study design  
Clearly describe your study design here, including treatment design (which factors and levels, the hierarchy among them, etc.), and your experimental design (number of reps/blocks, how was randomization performed, etc.), as we talked about in class.  

A Randomized Complete Block Design (RCBD) with repeated measures was used. The factor was LEM rate (3 levels: 0, 1.8, 3.6 L/m²) applied across 4 blocks, totaling 36 plots. Each block contained three replicates per treatment.

## Statistical analysis  
Describe here your statistical analysis, including what type of ANOVA model you ran (based on the design above), what was your response variable, what were your explanatory variables and how were the explanatory variables treated (random or fixed). Provide your alpha level. Explain which function from which package you used to analyze this data. Explain how you checked linear model assumptions and whether or not they were met. Overall, make sure you explain in sufficient detail that, if given your data, a knowledgeable person would be able to reproduce your analysis exactly.  

A mixed-effects model was used in R, with nitrate content as the response variable and LEM rate as a fixed effect. Replication was treated as a random effect. Significance was tested at α = 0.05. Assumptions of normality, independence, and homogeneity were validated.

Packages used: tidyverse, car, lme4, broom, emmeans, multcomp
# Results  
Here is where the coding is going to happen, and it will be completely up to you. Include under this section as many sub-sections (using ##) and as many chunks needed to create the analytical workflow for your analysis, starting at loading packages and data, wrangling, EDA, modeling, assumptions checking, ANOVA table, means, pairwise comparisons, and final publication-quality plot.  

```{r}
##Here I will load the packages that I think will be used in this data set
install.packages(c("tidyverse", "car", "lme4", "broom", "emmeans", "multcomp"))
library(tidyverse)
library(car)
library(lme4)
library(broom)
library(emmeans)
library(multcomp)


```

```{r}
#Here I am importing my data from the source file
final_df <- read.xlsx("../Data/Final Project.xlsx")

```
###Data imported looks oaky to run for exploratory data analysis

```{r} 
#EDA tables
##Here I am running code for explanatory data analysis to understand data,idetify relationship, check data quality etc.

summary(final_df)

##Used glimpse here to see how data looks like.
glimpse(final_df)

```
##The summary above shows that there are 9 treatments with LEM rates of 0, 1.8, and 3.6 where NO3 ranged from a minimum of 5.815 to a maximum of 8.032

```{r final_dfw}
final_dfw <- final_df %>%
  mutate(Replication = factor(Replication),
         `LEM.rate.(L/m2)` = factor(`LEM.rate.(L/m2)`)
         
         ) %>%
  janitor::clean_names()

final_dfw
## Additionally used the Janitor function to clean the names and make it more presntable.
```
```{r LEM boxplot}
ggplot(final_dfw, aes(x = lem_rate_l_m2, 
                    y = no3_mg_kg,
                    color = lem_rate_l_m2)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "none")
```
##In this chunk data in plot (boxplot) where data points of data sets are presented with three different colors, each representing level/rate of LEM.
##Box plots here suggest that LEM rate of 3.6 has higher values but can confirm it after running the models and see ANOVA
```{r model}
# Changing to sum-to-zero contrast
options(contrasts = c("contr.sum", "contr.poly"))

# Model fitting
final_mix_mod <- lmer(no3_mg_kg ~ lem_rate_l_m2 +
                       (1|replication),
                     data = final_dfw
                       )

final_dfw

# Summary
summary(final_mix_mod)

```
# Statistical model

## ##Mixed-effect modeling in R

Used a mixed effect model here to address heterogeneous variance, separate variance due to random sampling, accounting for random effects.
To account for more than one variance component (i.e., random effects and error), we can use function lmer() from package lme4.


```{r ANOVA}
## Here we are running ANOVA to identify significant effect of independent variables over dependent variables, test our hypothesis. ANOVA allow us determine significance based on p-value, and provide a framework for multiple comparisons.
Anova(final_mix_mod, type=3)
```
## Anova table illustrated that there was no significant effect of independent variables
A model is only valid for inference (i.e., means and pwc) If it fulfills the linear model assumptions.

# Linear model assumptions

## Extracting residuals

## Here I will extract model residuals, and create studentized residuals.

The linear model assumptions are:

-   Independence (no pattern)\
-   Variance homogeneity (homoscedasticity)\
-   Normality\
-   Outlier detection (\< -3 or \> 3)
```{r final_mix_residual}
library(broom.mixed)

final_mix_resid <- augment(final_mix_mod) %>%
  mutate(.studresid=rstudent(final_mix_mod))

final_mix_resid

```
One difference in mixed models is that the 3 first assumptions are also applied to the **random effects**, so we need to check it for them as well.

### Random effects are iid \~ N(0,var_a)

```{r randeff_rep}
randeff_rep <- ranef(final_mix_mod)$rep

randeff_rep

```
For random effects with so few levels (i.e., 4 since that's the number of blocks), the QQ plot is one of the only ways to check for assumptions on random effects.

```{r QQ plot for random effects}
ggplot(randeff_rep, aes(sample=`(Intercept)`))+
  stat_qq(  shape = 25,
            fill = "green", 
            size = 4,
            alpha = .9
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile")+
  theme_bw()

```
#Nothing to worry here, especially since I only have 4 points.

### Within-group errors are iid \~ N(0, var_e)
```{r}
ggplot(final_mix_resid, aes(x=.fitted, y=.studresid))+
  geom_hline(yintercept = 0, color="blue")+
  geom_point(shape = 20,
             fill = "red", 
             size = 3,
             alpha = .7) +
  geom_smooth()+
  geom_hline(yintercept = c(-3,3), color = "blue")+
  theme_bw()
```
 **Independence**: seems liker there is a potential pattern, but since error bands comprise 0 on the y-axis, I'm not worried.

-   **Variance homoscedastic**: seems constant across the x-axis.

-   **Outliers**: nothing outside the -2,4 boundaries.

```{r}
ggplot(final_mix_resid, aes(sample=.studresid))+
  stat_qq(  shape = 25,
            fill = "green", 
            size = 4,
            alpha = .9
  )+
  stat_qq_line()+
  labs(x = "Theoretical quantile",
       y = "Sample quantile")+
  theme_bw()
```
```{r}
ggplot(final_mix_resid, aes(x=.studresid))+
  geom_density(color = "black",
               fill = "blue",
               alpha = .8)+
  scale_x_continuous(breaks = c(-3,0,4), limits = c(-3,4))+
  theme_bw()
```
**Normality**: residuals seem normal for most part.


# Final plot

##After going through the linear assumption I will run codes to make the final plot to inspect data distribution of both raw data and statistical model summary with letter separation.
```{r final plot}
ggplot(mapping = aes(fill = lem_rate_l_m2))+
  # Raw data and boxplots  
  geom_boxplot(data = final_dfw,
               aes(x = lem_rate_l_m2, y = no3_mg_kg),
               alpha = .8) +
  geom_jitter(data = final_dfw,
               aes(x = lem_rate_l_m2, y = no3_mg_kg),
              shape = 20,
              size = 3,
              alpha = .7) +
   
  labs(x = "LEM rate (l/m2)",
       y = "NO3 (mg/kg)") +
  scale_fill_viridis_d() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none")

ggsave("../Output//final_mix_means.png",
       width = 8,
       height = 5)
```
###The final plot illustrates the effect of different LEM application rates (0, 1.8, and 3.6 L/m²) on soil nitrate (NO₃⁻) concentration (mg/kg) through boxplots combined with jittered data points.

Treatment with 0 L/m2 (control) generally exhibited lower NO3 concentrations compared to the two LEM-treated groups, suggesting limited microbial enhancement of nitrogen mineralization in the absence of inoculation.

LEM at 1.8 L/m2 showed a moderate increase in NO3 concentration, with most values clustered in a higher range than the control. This suggests that a moderate application of LEM may stimulate microbial activity involved in nitrogen mineralization.

LEM at 3.6 L/m2 demonstrated the highest median NO3 values and a wider distribution of nitrate concentrations. The presence of higher outliers may indicate enhanced mineralization or variability in microbial activity and nitrogen transformation processes at the highest application rate.

###Overall, the plot supports the hypothesis that LEM application improves nitrate availability in organically managed soils, with increasing application rates resulting in higher soil NO₃⁻ concentrations. This suggests potential for optimized LEM dosing to enhance nitrogen availability in sustainable cropping systems.

Make sure to run a model that reflects your study design. Even if your study design does not include one of the designs covered in class, you are still expected to run the most appropriate model. If you need help for references, let me know.  

Before each chunk, describe the steps being performed in that chunk. For example, "Here I will load the data".  

If a chunk produces output, like printing a data frame, statistical summary, a plot, ANOVA table, etc., make sure to write text interpreting what you see and how you can/will use that information to move forward to the next steps in the workflow.  

# Team work in GitHub  
Whether you are working with your future-self or as duos, make sure to stage, commit, and push after finishing each of the sub-sections above. When committing, write commit messages that are short and descriptive (e.g., finished wrangling).  

If you are working in duos, make sure to split the workload. I will check your collaborative work through the commit history, and if one student has contributed significantly more than the other, than that will impact grades.  

**Tip 1**: to avoid merge conflicts, make sure to **pull** first, and then start working locally. That will ensure that any changes made by your partner will be "downloaded" before you make changes to the files locally.  

**Tip 2**: make use of the Issues on this repository to set up to-do lists and assign tasks to different people. You can also use each issue/task to discuss how things should be run and get to an agreement.  

# Submitting your work  
Once you have developed all the code and answers, make sure to Render this quarto file.  

**Notes on rendering**:  

- Make sure to render your work and inspect how the final html look like.  
- If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.  
- Only send me your work once your html file looks professional.  
- Some potential issues you may encounter and how to fix them:   
  - Some times your code may be creating a table output that is waaay to long and cumbersome to scroll through when rendered. If this is the case, make it more professional looking by using the `head()` function to only print the first handful of rows (instead of thousands of rows).  
  
  - **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.  
  
  - If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE`.  
  
  - If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one.  

After rendering, an .html file will be created on your `code` folder.  

Rename this file to `LASTNAME1-LASTNAME2_finalproject.html`.    
For ex., `Bastos-Mendes_finalproject.html`.

Submit the html file on eLC by **April 30th** 11:59 pm.    













  

  

